---
title: "Hepatitis Education & Outreach"
author: Samuel Muma, Adharsh Maheswaran, Chantel Orimoloye, Mikhail Tabong, Omodara
  Ojebuoboh
date: "`r Sys.Date()`"
output: html_document
---

# Importing dataset

```{r}

# Required packages
required_packages <- c("readr", "dplyr", "tidyr", "ggplot2", "corrplot", "RColorBrewer", "caret", "knn", "e1071") 

new_packages <- required_packages[!required_packages %in% installed.packages()]
if (length(new_packages) > 0) {
  install.packages(new_packages)
}

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(corrplot)
library(RColorBrewer)
library(caret)
library(kknn)
library(e1071)

# Read in CSV
hepatitis_df <- read_csv("hepatitis.csv")
head(hepatitis_df)

```

# 1. Data Preprocessing

a. Handle missing values appropriately

```{r}
# Getting the most frequent value (mode) for binary categorical columns
binary_columns <- c("Steroid", "Liver Big", "Liver Firm", 
                    "Spleen Palpable", "Spiders", "Ascites", "Varices")

# Helper funciton to impute missing values with the mode
impute_mode <- function(col) {
  mode_value <- as.numeric(names(sort(table(col), decreasing = TRUE))[1])
  col[is.na(col)] <- mode_value
  return(col)
}

# Apply imputation
hepatitis_df[binary_columns] <- lapply(hepatitis_df[binary_columns], impute_mode)

# Handling Nominal columns
mean_to_impute <- c("Bilirubin", "Alk Phosphate", "Sgot")
median_to_impute <- c("Albumin", "Protime")

# Helper function to impute missing values with the mean or median
impute_with <- function(data, columns, method = "mean") {
  for (col in columns) { # Fix: Added parentheses around "col in columns"
    if (method == "mean") {
      impute_value <- mean(data[[col]], na.rm = TRUE)
    } else if (method == "median") {
      impute_value <- median(data[[col]], na.rm = TRUE) 
    } else {
      stop("Invalid method. Please use 'mean' or 'median'.")
    }
    # Impute N/A in the specified column
    data[[col]][is.na(data[[col]])] <- impute_value
  }
  return(data)
}

# Impute using mean & median
hepatitis_df <- impute_with(hepatitis_df, mean_to_impute, method = "mean")
hepatitis_df <- impute_with(hepatitis_df, median_to_impute, method = "median")
head(hepatitis_df)

```

b. Encode categorical variables as needed

```{r}
# Columns to re-code (1 = No (except Sex where Sex = Male & Class 1 = Die), 2 = No (except Sex where Sex = Female & Class 2 = Yes))
features_to_recode <- c("Sex", "Steroid", "Antivirals", "Fatigue",
               "Malaise","Anorexia", "Liver Big", "Liver Firm", 
               "Spleen Palpable", "Spiders", "Ascites", "Varices",
               "Histology", "Class")

# Function to recode 1's to 1's and 2's to 0's
recode_binary <- function(column, column_name) {
  if (column_name == "Sex") {
    #For Sex: 1 = Male, 2 = Female
    return(ifelse(column == 2, 0, 1)) # 1 = Male, 0 = Female
  } else if (column_name == "Class") {
    # For Class: 1 = Die, 2 = Live
    return(ifelse(column == 2, 1, 0)) # 1 = Live, 0 = Die
  } else {
    # For other columns: 1 = No, 2 = Yes
    return(ifelse(column == 2, 1, 0)) # 1 = Yes, 0 = No
  }
}

# Applying the re-coding function
for (col in to_recode) {
  hepatitis_df[[col]] <- recode_binary(hepatitis_df[[col]], col)
}

# Convert 'Class' to a factor
hepatitis_df$Class <- factor(hepatitis_df$Class, levels = c(0, 1), labels = c("Die", "Live"))
```

c. Normalize or standardize features if required

```{r}
# Checking summary statistics for standardization
summary(hepatitis_df$Bilirubin)
summary(hepatitis_df$`Alk Phosphate`)
summary(hepatitis_df$Sgot)
summary(hepatitis_df$Protime)

# Standardizing necessary features
features_to_scale <- c("Bilirubin", "Alk Phosphate", "Sgot", "Protime")
hepatitis_df[features_to_scale] <- scale(hepatitis_df[features_to_scale])
summary(hepatitis_df[features_to_scale])
```

# 2. Data Visualization

a. Create at least 3 meaningful visualizations to explore data distributions and relationships.

```{r}
# Setting a colour palette
palette <- brewer.pal(3, "Set2")

# --- Histograms -- #
# 1. Histogram for Bilirubin
ggplot(hepatitis_df, aes(x = Bilirubin)) +
  geom_histogram(binwidth = 0.5, fill = palette[1], color = "black", alpha = 0.7) + 
  labs(title = "Histogram of Bilirubin", x = "Bilirubin", y = "Frequency") +
  theme_classic() + 
  theme(text = element_text(size= 12))

# 2. Histogram for Alk Phosphate
ggplot(hepatitis_df, aes(x = `Alk Phosphate`)) +
  geom_histogram(binwidth = 0.5, fill = palette[2], color = "black", alpha = 0.7) + 
  labs(title = "Histogram of Alk Phosphate", x = "Alk Phosphate", y = "Frequency") + 
  theme_classic() + 
  theme(text = element_text(size= 12))

# 3. Histogram for Sgot
ggplot(hepatitis_df, aes(x = Sgot)) +
  geom_histogram(binwidth = 0.5, fill = palette[3], color = "black", alpha = 0.7) + 
  labs(title = "Histogram of Sgot", x = "Sgot", y = "Frequency") +
  theme_classic() + theme(text = element_text(size= 12))

# --- Boxplots --- #
#1. Boxplot for Bilirubin by Class
ggplot(hepatitis_df, aes(x = factor(Class), y = Bilirubin, fill = factor(Class))) +
  geom_boxplot() + labs(title = "Boxplot for Bilirubin by Outcome (Class)", x = "Outcome", y = "Bilirubin") + 
  theme_classic() +
  scale_fill_manual(values = palette) + 
  theme(text = element_text(size = 12))

#2. Boxplot for Alk Phosphate by Class
ggplot(hepatitis_df, aes(x = factor(Class), y = `Alk Phosphate`, fill = factor(Class))) +
  geom_boxplot() + labs(title = "Boxplot for Alk Phosphate by Outcome (Class)", x = "Outcome", y = "Alk Phosphate") + 
  theme_classic() +
  scale_fill_manual(values = palette) + 
  theme(text = element_text(size = 12))

#3. Boxplot for Sgot by Class,
ggplot(hepatitis_df, aes(x = factor(Class), y = Sgot, fill = factor(Class))) +
  geom_boxplot() + labs(title = "Boxplot for Sgot by Outcome (Class)", x = "Outcome", y = "Sgot") + 
  theme_classic() +
  scale_fill_manual(values = palette) + 
  theme(text = element_text(size = 12))

# --- Correlation Heatmap --- #
hepatitis_df_without_target <- hepatitis_df %>% select(-Class)
numeric_features <- sapply(hepatitis_df_without_target, is.numeric)
cor_matrix <- cor(hepatitis_df_without_target[, numeric_features], use = "complete.obs")
corrplot(cor_matrix, method = "number", type = "upper",
         tl.col = "black", tl.cex = 0.8, addCoef.col = "black",
         insig = "blank",
         sig.level = 0.05,
         title = "Correlation Heatmap of All Features",
         mar = c(0, 0, 1, 0), col = rev(brewer.pal(9, "Blues")))
```

# 3. Model Building

a. Split the dataset into 75% training and 25% testing sests

```{r}
#set seed for reproducibility
set.seed(123)

# Splitting dataset
training_df <- createDataPartition(hepatitis_df$Class, p = 0.75, list = FALSE)
train <- hepatitis_df[training_df, ]
test <- hepatitis_df[-training_df, ]

train$Class <- as.factor(train$Class)

# Cross-validation with 5 folds and configure ROC-based evaluation
ctrl <- trainControl(method = "cv",
                     number = 5,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)

# Removing NA variables in the training dataset
calculate_mode <- function(x) {
  unique_x <- unique(na.omit(x))
  unique_x[which.max(tabulate(match(x, unique_x)))]
}

for (col in colnames(train)) {
  if (any(is.na(train[[col]]))) {
    mode_value <- calculate_mode(train[[col]])
    train[[col]][is.na(train[[col]])] <- mode_value
  }
}

# Checking if there are any missing values
sum(is.na(train))

# Removing spaces in column names
colnames(train) <- gsub(" ", "_", colnames(train))

# Fixing Class levels
levels(train$Class)  # Check current levels
train$Class <- relevel(train$Class, ref = "Live") 

# Making the data frame type the standard data.frame
train <- as.data.frame(train)

```

b. Training a Decision Tree, SVM, and KNN

```{r}
# Decision Tree
Decision_Tree <- train(Class ~ ., data = train,
                       method = "rpart",
                       trControl = ctrl,
                       metric = "ROC")

# SVM
SVM <- train(Class ~ ., data = train,
             method = "svmRadial",
             trControl = ctrl, 
             metric = "ROC")

# KNN
KNN <- train(Class ~ ., data = train,
                   method = "kknn",
                   trControl = ctrl,
                   metric = "ROC")
```

# 4. Cross-Validation Model Evaluation